---
title: "p8105_hw3_mp3745"
author: "Matthew Parker"
date: '`r format(Sys.time(), "%Y-%m-%d")`'
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(viridis)

knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	fig.width = 8, 
  fig.height = 6,
  out.width = "90%"
)
options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)
scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
theme_set(theme_minimal() + theme(legend.position = "bottom"))
```


## Problem 1

#### 1)

Enter in data
```{r}
cancer = tibble(
  age = rep(c(25, 35, 45, 55, 65, 75), 2),
  alc_cons = c(rep("0-79", 6), rep("80+", 6)),
  case = c(0, 5, 21, 34, 36, 8, 1, 4, 25, 42, 19, 5),
  ctrl = c(106, 164, 138, 139, 88, 31, 9, 26, 29, 27, 18, 0)
)
```

Model
```{r}
# Response matrix
resp = cancer %>% 
  select(case, ctrl) %>% 
  as.matrix()

# Model
logit_cancer = glm(resp ~ age + alc_cons, family = binomial(link = 'logit'), data = cancer)

logit_cancer$coefficients[1]

logit_cancer %>% 
  broom::tidy() %>% 
  knitr::kable()
```

From the logit model, the log odds of esophageal cancer for someone with daily alcohol consumption of 0-79g that is age 0 is `r logit_cancer$coefficients[1]`. The log odds ratio of esophageal cancer comparing someone with daily alcohol consumption 80+g to 0-79g is `r logit_cancer$coefficients[3]`, holding age constant. And the log odds ratio of esophageal cancer associated with a one year increase in age is `r logit_cancer$coefficients[2]`, holding daily alcohol consumption constant. From these results, age and is positively associated with esophageal cancer and a higher daily alcohol consumption (80+g) is also associated with esophageal cancer.

