---
title: "p8105_hw3_mp3745"
author: "Matthew Parker"
date: '`r format(Sys.time(), "%Y-%m-%d")`'
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(viridis)

knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	fig.width = 8, 
  fig.height = 6,
  out.width = "90%"
)
options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)
scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
theme_set(theme_minimal() + theme(legend.position = "bottom"))
```


## Problem 1

#### 1)

Enter in data
```{r}
cancer = tibble(
  age = rep(c(25, 35, 45, 55, 65, 75), 2),
  alc_cons = c(rep("0-79", 6), rep("80+", 6)),
  case = c(0, 5, 21, 34, 36, 8, 1, 4, 25, 42, 19, 5),
  ctrl = c(106, 164, 138, 139, 88, 31, 9, 26, 29, 27, 18, 0)
)
```

Model
```{r}
# Response matrix
resp = cancer %>% 
  select(case, ctrl) %>% 
  as.matrix()

# Model
logit_cancer = glm(resp ~ age + alc_cons, family = binomial(link = 'logit'), data = cancer)

logit_cancer %>% 
  broom::tidy() %>% 
  knitr::kable()
```

From the logit model, the log odds of esophageal cancer for someone with daily alcohol consumption of 0-79g that is age 0 is `r logit_cancer$coefficients[1]`. The log odds ratio of esophageal cancer comparing someone with daily alcohol consumption 80+g to 0-79g is `r logit_cancer$coefficients[3]`, holding age constant. And the log odds ratio of esophageal cancer associated with a one year increase in age is `r logit_cancer$coefficients[2]`, holding daily alcohol consumption constant. From these results, age and is positively associated with esophageal cancer and a higher daily alcohol consumption (80+g) is also associated with esophageal cancer.



## Problem 2

#### 1)

Enter in data
```{r}
germ = tibble(
  species = c("oa_75", "oa_75", "oa_73", "oa_73"),
  root = c("bean", "cucumber", "bean", "cucumber"),
  germinated = c(10 + 23 + 23 + 26 + 17, 5 + 53 + 55 + 32 + 46 + 10, 8 + 10 + 8 + 23 + 0, 3 + 22 + 15 + 32 + 3),
  total = c(39 + 62 + 81 + 51 + 39, 6 + 74 + 72 + 51 + 79 + 13, 16 + 30 + 28 + 45 + 4, 12 + 41 + 30 + 51 + 7),
  failed = total - germinated
)

# response
germ_resp = germ %>% 
  select(germinated, failed) %>% 
  as.matrix()
```

Fit model
```{r}
germ_logit = glm(germ_resp ~ species + root, family = binomial(link = 'logit'), data = germ)

germ_logit %>% 
  broom::tidy() %>% 
  knitr::kable()
```

From the logit model, the log odds of germinating for an O. aegyptiaca 73 seed with bean root extract is `r germ_logit$coefficients[1]`. The log odds ratio of germinating comparing an O. aegyptiaca 75 seed with an O. aegyptiaca 73 seed is `r germ_logit$coefficients[2]`, holding root extract media constant. The log odds ratio of germinating comparing a seed with cucumber root extract to bean root extract is `r germ_logit$coefficients[3]`, holding seed type constant. From these results, root extract medium is strongly associated with germinating and O. aegyptiaca 75 is also associated with germinating.


#### 2)

Check for lack of fit
```{r}
pval = 1 - pchisq(germ_logit$deviance, 4 - 3)
```

Since p-value = `r pval` < 0.05, we have sufficient evidence to reject the null hypothesis that the model fits the data well.


Check for overdispersion
```{r}
res = residuals(germ_logit, type = 'pearson')

resid_points = tibble(
  x = qnorm((4 + 1:4 + 0.5)/(2 * 4 + 1.125)),
  y = sort(abs(res))
)

resid_points %>% 
  ggplot(aes(x = x, y = y)) +
  geom_point() + 
  geom_abline(slope = 1) + 
  scale_x_continuous(limits = c(0, 2)) +
  scale_y_continuous(limits = c(0, 2)) +
  labs(
    x = "Expected Half-Normal Order Stats",
    y = "Ordered Abs Pearson Residuals"
  )
```


Based on above plot, it appears they may be a liner deviation from the reference line, which indicates possible overdispersion.

Calculate estimate of dispersion parameter
```{r}
g_stat = sum(residuals(germ_logit, type = 'pearson')^2)

phi = g_stat/(4 - 3)
```

The estimated dispersion parameter is `r phi`

Update model with dispersion parameter
```{r}
summary(germ_logit, dispersion = phi)
```

In the updated model, our estimates and their interpretations are still valid. In the updated model, the std. error has been inflated by `r phi`. This has caused the z value and Pr(>|z|) to change. The coefficient for the term comparing species of seed in the model has now become insignificant, indicating there may not be a significant association between seed species and germinating when accounting for root extract media.